Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity

### 2022-02-14 docs/manuscript_v8.Rmd
some ms and prepping for final validation

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via gas exchange. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$, the rate of this diffusion. We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established Bayesian inference techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (SWOT has not yet launched) and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Like similar Bayesian remote sensing algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$rBIAS), 2)*100`%). Finally, we show BIKER robust to measurement errors implicit in the SWOT data. With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and water turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [*Liu accepted*; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' incredibly small percentage of the global land surface [0.67%- *Liu accepted*], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

@@

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
There is a robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a]. This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthlytemporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking. As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.

@@

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what *k* must have been to produce their observations. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity. Even though approaches like this are incredibly useful for expanding our mechanistic understanding of gas exchange, it is less useful for global upscaling purposes as it relies on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].

@@

To answer this first question, we use hydraulic geometry [the fundamental geomorphic relationships between streamflow and channel shape- @leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined a wider than 50m) gas exchange. We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) under both a 'no-measurement-error' and a 'measurement-error' scenario to explore algorithm robustness to the error tolerances expected in the satellite's measurements. Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that are reliant on in situ hydraulic measurements.

@@

## 2 Prediciting $k_{600}$ from large-river hydraulic geometry
To predict $k_{600}$ in the large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a suite of models we empirically test. Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that are compatible with SWOT measurements (section 2.2). We then validate the model (section 2.3)

@@

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. This is equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface. Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$
Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall via equation 5 [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993]. Another approach models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that equation 6l can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

@@

### 2.3 Deriving a large-river $k_{600}$ model
Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$. This is a common assumption in hydraulic and geomorphic modeling of large rivers, and the average flow in a SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

We therefore assume that all SWOT-observable rivers are hydraulically-wide. We then test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and equations 5-6 outlined above. The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT. While the complete model derivations and results for the four models are provided in Text S2 and Figure S1, the final and best-performing model is presented below.

We impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$). Equation 7 thus define gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.

$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(7)}$$
### 2.4 Model validation
With equation 7 derived, we now empirically its strength of fit for hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was operationally defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with. The model is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2. Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

@@

Our model explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well. While the individual residuals can be quite large, the overall scaling of $k_{600}$ with river hydraulics is strongly captured. Compared to the other models tested (Figure S1), there is also less bias in the estimates thanks to the addition of the Reynold's number scaling for low-turbulent flows. The success of this model in hydraulically-wide channels provides us with a strong physical-model for gas evasion. The river hydraulics terms in equation 7 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity. This is explored next.

## 3 Exploiting equation 8 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers. Further, equation 8 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, chanel depth and velocity from SWOT observations using a modified form of Manning's equation. Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).

### 3.1 Data
![Figure 3: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced.](cache/maps/map3.jpg)

To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch. In the hydrology literature, it is therefore standard practice to benchmark SWOT-related algorithms on "SWOT-like" data [@durandIntercomparisonRemoteSensing2016]. We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER. These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016 as benchmarking datasets to explore remote sensing of discharge (RSQ) algorithm performance for the SWOT mission. These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers. These `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers represent the 'no-measurement-error' scenario that reflects a best-case scenario that is used here as a first validation of BIKER. `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then further prepared by @frassonExploringFactorsControlling2021 to more closely mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track. Specifically, they account for 1) the SWOT error budget itself [CITE}], 2) thermal noise, and 3) radar layover error, witht the latter two being modeled by [@durandHowWillRadar2020].

*COLIN: @frassonExploringFactorsControlling2021 also developed a sophisticated error model that mimics the effects of the atmosphere, topography, and water surface geometry on SWOT measurements (16/47 rivers) to add realistic radar layover errors that depend on the time and place of observation. The resulting data are the most realistic synthetic SWOT river data in the literature. The approximate locations of these rivers are plotted in Figure 1.*

**According to Durand 2016 and Frasson, there is no measurement error added to the non-error scenario rivers. Mike had a small bit of Gaussian noise to pepsi 1 but he doesn't seem to count it has true measurement error.**

For the $CO_2$ evasion and carbon emissions calculations (section 3.1.3), we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River). Therefore, the raw carbon emissions estimates presented in this paper are meaningless in the context of actually measured carbon emissions from these rivers. However, we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes, and therefore applying these 'unit' $CO_2$ values allows such a compariosn by providing a realtistic timeseries.

### 3.2 Section 3 methods
#### 3.2.1 BIKER
BIKER, and Bayesian inference in general, starts from Bayes rule (equation 8), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters. Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution. Note that $p(x)$ is usually computationally intractable to integrate exactly, but Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$. Sampling algorithms are then used to approximate the posterior distribution, as is done in BIKER.

>>>>

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via gas exchange. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established Bayesian inference techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (SWOT has not yet launched) and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Like similar Bayesian remote sensing algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$rBIAS), 2)*100`%). Finally, we show BIKER robust to measurement errors implicit in the SWOT data. With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and near-surface turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [*Liu accepted*; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' incredibly small percentage of the global land surface [0.67%- *Liu accepted*], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

@@

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
There is a robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a]. This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking. As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.

@@

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what *k* must have been to produce their 'observations'. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity. Even though approaches like this are incredibly useful for expanding our mechanistic understanding of gas exchange, it is less useful for global upscaling purposes as it relies on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].

@@

To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m) gas exchange. We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) under two 'measurement-error' scenarios to explore algorithm robustness to the expected measurement errors implicit in the satellite's observations. Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that are reliant on in situ hydraulic measurements.

@@

## 2 Prediciting $k_{600}$ from large-river hydraulic geometry
To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test. Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at an equation that is compatible with SWOT measurements (section 2.3). We then empirically validate the model (section 2.3)

@@

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. This is equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence scenarios where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface. Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$
Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall [equation 5- @lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993]. Another approach (specific to fluvial settings) models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that equation 6 can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

@@

### 2.3 Deriving a large-river $k_{600}$ model
Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$. This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average flow in a SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

We therefore assume that all SWOT-observable rivers are hydraulically-wide and derive a model for gas exchange. The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT. To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$). Equation 7 thus define gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.

$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(7)}$$

we also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of equations 3-4 and equations 5-6. T While the complete model derivations and results for the four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented below.

### 2.4 Model validation
With equation 7 derived, we now empirically its strength of fit for hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was operationally defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with. The model is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2. Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

@@

Our model explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well. While the individual residuals can be quite large, the overall scaling of $k_{600}$ with river hydraulics is strongly captured. Compared to the other models tested (Figure S1), there is also less bias in the estimates thanks to the addition of the Reynold's number scaling for low-turbulent flows. The success of this model in hydraulically-wide channels provides us with a strong physical-model for gas evasion. The river hydraulics terms in equation 7 ($\bar{U}$, $H$, and $S$) can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity. This is explored next.

## 3 Exploiting equation 7 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers. Further, equation 7 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel depth and velocity from SWOT observations using a modified form of Manning's equation. Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).

### 3.1 Data
![Figure 3: Map of the 47 hydraulic models and 1 timeseries of $CO_2$ samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced.](cache/maps/map3.jpg)

To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch. In the hydrology literature, it is therefore standard practice to benchmark SWOT-related algorithms on "SWOT-like" data [@durandIntercomparisonRemoteSensing2016]. We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER. These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016 as benchmarking datasets to explore remote sensing of discharge (RSQ) algorithm performance for the SWOT mission. These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers.

Recall that we validate BIKER under two different 'error scenarios'. While SWOT will provide river surface measurements of novel quality and resolution, as with all remote sensing products there are expected errors that will be implicit in these measurements. Here, we validate BIKER under a 'no-measurement-error' scenario that reflects a best-case situation and is used as a first test of algorithm validity. `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then additionally validated under a 'measurement-error' scenario to more closely mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track. SWOT river error modeling was developed by @durandHowWillRadar2020 and then applied to these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers by @frassonExploringFactorsControlling2021.

For the $CO_2$ evasion and carbon emissions calculations (section 3.1.3), we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River). Therefore, the raw carbon emissions estimates presented in this paper are meaningless in the context of actually measured carbon emissions from these rivers. However, we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes, and therefore applying these 'unit' $CO_2$ values allows for such a comparison by providing a realistic timeseries.

### 3.2 Section 3 methods
#### 3.2.1 BIKER
**You made a potential note about moving this to the SI**
BIKER, and Bayesian inference in general, starts from Bayes rule (equation 8), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters. Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution. Note that $p(x)$ is usually computationally intractable to integrate exactly, but Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$. Sampling algorithms are then used to approximate the posterior distribution, as is done in BIKER.

$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(8)}$$
The heart of BIKER is its reformulation of the $k_{600}$ model (equation 7) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e. $f(x|\Theta)$). This is similar to the 'McFLI' ('Mass-Conserved Flow Law Inversion') logic used in some SWOT RSQ algorithms [@gleasonTrackingRiverFlows2017]. To start, we write $k_{600}$ as a function of SWOT-observables $W$ and $S$. This algebra is carried out using equation 7, the fitted value for $\beta_1$ from Figure 2 (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$). Following section 2.3, we continue to assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$). To leverage additional SWOT data, the wetted channel area *A* is further split into the the SWOT-unobservable portion $A_0$ and SWOT-observable portion $dA$ following @durandEstimatingReachaveragedDischarge2014 and @hagemannBAMBayesianAMHGManning2017 where $dA_{it} = \sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_{t'}}$ for cross-section *i* and timestep *t*.
